---
title: "fraud_detection"
author: "Akoua Orsot"
date: "2/8/2022"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Fraud Detection

In this notebook, we will attempt to build an algorithm able to detect a fraudulent transaction using a training dataset. We will exlain the thinking process at every step usine LIME (Local Interpretable Model-agnostic Explanations) principles making it accessible and user-friendly.

## 1. Environment Set-up

### a) Importing libraries

```{r}

library(tidyverse)
library(ggplot2)
library(e1071)
library(dplyr)
```

### b) Loading dataset

```{r}
df <- read_csv(file = 'C:/Users/Akoua Orsot/Desktop/ds_projects_data/creditcard.csv')
head(df)
```

## 2. Initial Diagnostics

### a) Glimpse of the data

```{r}
df %>% str()
```

### b) Descriptive Statistics

```{r}
df %>% summary()
```

**Takeaway**

It confirms the note in the project description; indeed, we have a considerable class imbalance with the target variable. It stays consistent with the fact that most fraudulent activities are much less frequenct than non-fraudulent. We shall take note of it before proceeding in order to avoid any overfitting issue when fitting the machine learning models.

### c) Target Variable Analysis

```{r}
  df %>% group_by(Class) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 5)) %>% 
  arrange(desc(freq))
```

**Note**

For privacy purposes, we did not have any information on the numerical predictors as those were transformed with PCA except Amount & Time. In that regard, Amount presented itself as potentially most informative for the feature variable analysis.

With an onslaught of outliers, we had to transform it using log scale to obtain a better view of the variable's distribution

```{r}
df$Amount %>% summary()
```

```{r}
df %>% ggplot(aes(Amount)) +
  geom_histogram(bins=35) +
  scale_x_log10() +
  labs(
  x = "Dollar Amount (Log Scale)",
  y = "Frequency (Count)",
  title= "Distribution of Transaction Amount (log scaled)"
 )
```
